---
output: github_document
---

# aSCAC (*Archive des styles de céramique en Afrique centrale*)

**Dirk Seidensticker**

```{r setup, include=FALSE}
library(concaveman)
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(sf)
library(rnaturalearth)
library(viridis)

land10 <- ne_download(scale = 10, type = "land", category = "physical", returnclass="sf")
rivers10 <- ne_download(scale = 10, type = "rivers_lake_centerlines", category = "physical", returnclass = "sf")
lakes10 <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass="sf")

sites <- data.table::fread("sites.csv", encoding = "UTF-8")  %>%
  st_as_sf(crs = 4326, 
           coords = c("LONG", 
                      "LAT"), 
           remove = FALSE, 
           na.fail = F)

pottery <- data.table::fread("potterygroups.csv", encoding = "UTF-8")

sites.meta <- sites %>% 
  dplyr::left_join(pottery, by = "POTTERY")
```

This repository contains an archive of pottery groups in Central Africa.

The initial version of the data was first published in:

Seidensticker, D., W. Hubau, D. Verschuren, C. Fortes-Lima, P. de Maret, C.M. Schlebusch & K. Bostoen. 2021. Population Collapse in Congo Rainforest from AD 400 Urges Reassessment of the Bantu Expansion. Science Advances 7. https://doi.org/10.1126/sciadv.abd8352. (data: https://github.com/dirkseidensticker/HumActCentralAfrica_Paper)


```{r analysis, include=FALSE}
breaks <- seq(-1000, 2000, 100)
class <- seq(1,length(breaks), 1)
breaks <- data.frame(breaks, class)
for(i in 1:nrow(breaks)){
  breaks[i, "labels"] <- paste0(breaks[i,"class"], ": ", breaks[i,"breaks"], "/", breaks[i+1,"breaks"])
}

# Frequency of sites per pottery group

pottery.sites.freq <- as.data.frame(stats::aggregate(SITE ~ POTTERY, 
                                                     data =sites, 
                                                     FUN = length)) 

# Area per pottery group (Convex hull)
# see https://github.com/joelgombin/concaveman

id <- dplyr::filter(pottery.sites.freq, SITE > 2)

pottery.sites.area <- sf::st_multipolygon()
pottery.sites.area <- st_sf(polygons = st_sfc(st_polygon()))
sf::st_crs(pottery.sites.area) <- (4326)
pottery.sites.area$POTTERY <- NA
for(i in 1:nrow(id)){
  sites.f <- dplyr::filter(sites, POTTERY == id[i,1])
  conc.hull <- concaveman(sites.f)
  conc.hull$POTTERY = id[i, "POTTERY"]
  pottery.sites.area <- rbind(pottery.sites.area, conc.hull)
}

pottery.sites.area$AREA <- sf::st_area(pottery.sites.area)
pottery.sites.area$AREA <- as.numeric(pottery.sites.area$AREA)/1E9 # convert m2 into k(ilo) km2

pottery.sites.area <- pottery.sites.area %>%
  dplyr::left_join(pottery, by = "POTTERY")

# Frequency of pottery groups per bin

pottery.cent <- data.frame(matrix(ncol = ncol(pottery)+1, nrow = 0))
x <- c(names(pottery), "CLASS")
colnames(pottery.cent) <- x

for (i in 1:length(pottery$POTTERY)){
  for (j in 1:(nrow(breaks)-1)) {
    if(pottery[i,"TO"] > breaks[j,"breaks"] & 
       pottery[i,"FROM"] < breaks[j+1,"breaks"]){
      l <- pottery[i,]
      l$CLASS <- breaks[j,"labels"]
      pottery.cent <- rbind(pottery.cent, as.data.frame(l))
    }

  }
}
pottery.cent$AGE <- (as.numeric(sub("/.*", "", sub(".*? ", "", pottery.cent$CLASS))) + as.numeric(sub(".*/", "", sub(".*? ", "", pottery.cent$CLASS)))) / 2
pottery.cent$AGE.jitter 	<- jitter(pottery.cent$AGE, 2)

# Frequency of pottery groups per 100 years

pottery.cent.freq <- as.data.frame(table(pottery.cent$AGE))
pottery.cent.freq$Var1 <- as.numeric(as.character(pottery.cent.freq$Var1))

# merge into meta tables

pottery.cent.meta <- pottery.cent %>%
  dplyr::select(-DESCRIPTION) %>%
  dplyr::left_join(pottery.sites.freq, by = "POTTERY") %>%
  dplyr::left_join(pottery.sites.area, by = "POTTERY")

sites.cent <- merge(x = sites, 							# merge sites per style with class (200-year century list)
                    y = dplyr::select(pottery.cent, -DESCRIPTION), 
                    by = "POTTERY", 
                    allow.cartesian = TRUE)
```

Some basic statistics (see Fig. 3 in [Seidensticker et al. 2021](https://advances.sciencemag.org/content/7/7/eabd8352))

```{r stats, warning = FALSE, echo = FALSE, fig.align = 'center', dpi = 300, out.width = '75%'}
freq.plt <- ggplot() + 
  geom_bar(data = pottery.cent.freq, 
           aes(x = Var1, 
               weight = Freq), 
           fill = "white", 
           color = "#333333", 
           width = 75) + 
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_continuous("Number of  \n pottery groups", 
                     expand = c(0, 0)) + 
  theme_classic() + 
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())

qty.sites.plt <- ggplot() + 
  geom_boxplot(data = pottery.cent.meta, 
               aes(x = AGE, 
                   y = SITE, 
                   group = AGE), 
               outlier.shape = 3, 
               width = 75) + 
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_sqrt("Number of sites\n per pottery group", 
               expand = c(0, 0)) + 
  theme_classic() + 
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())

area.plt <- ggplot() + 
  geom_boxplot(data = pottery.cent.meta, 
               aes(x = AGE, 
                   y = AREA, 
                   group = AGE), 
               outlier.shape = 3, 
               width = 75) + 
  scale_x_continuous("cal BCE/CE", 
                     expand = c(0, 0), 
                     breaks = seq(-1000, 2000, 200)) + 
  scale_y_sqrt("Distribution area of \n pottery groups (1000 km^2)", 
               expand = c(0, 0)) + 
  theme_classic()

cowplot::plot_grid(freq.plt, 
                   qty.sites.plt, 
                   area.plt, 
                   ncol = 1, 
                   align = "v", axis = "lr", 
                   labels = "auto", 
                   rel_heights = c(1, 1, 1.5))
```

Map of the distribution areas of pottery styles through time:

```{r map, warning = FALSE, echo = FALSE, fig.align = 'center', dpi = 300, out.width = '100%'}
ggplot() + 
  geom_sf(data = land10, fill = "#d3d3d3") + 
  geom_sf(data = rivers10, size = .5, color = "grey") + 
  geom_sf(data = lakes10, fill = 'grey', color = NA) + 
  geom_sf(data = pottery.cent.meta %>% sf::st_as_sf(), 
          aes(fill = AGE)) + 
  scale_fill_viridis() + 
  facet_wrap(AGE~.) + 
  coord_sf(xlim = c(8, 27), 
           ylim = c(-10, 6)) + 
  theme_few()
```

## Datasets

> TODO

### sites.csv

| Datafield | Description |
| --------- | ----------- |



### potterygroups.csv

| Datafield | Description |
| --------- | ----------- |
